\Large \bf Vigor: Toward Carrier-Grade Software Network Functions

### 2016-11-22 doc/paper/paper.tex
bugfix in the example original code snippet (thanks, Luis!), line-number references.

\lstset{language=C,
    basicstyle=\footnotesize\ttfamily,
    keywordstyle=\color{myblue}\ttfamily,
    stringstyle=\color{mygreen}\ttfamily,
    commentstyle=\color{gray}\ttfamily,
    keepspaces=true,
    tabsize=2,
    otherkeywords={bool},
    numbers=left,
    numbersep=1pt,
    numberstyle=\tiny\color{gray}
}

@@

This section motivates and explains at a high level what it's like to use Vigor
when developing NFs. For the sake of illustration, say we wish to develop a
trivial stateful NF that implements the discard protocol~\cite{rfc863}: an
infinite loop receives packets from one interface, discards the ones that target
port 9 (``discard port''), and forwards the others through another interface.

@@

int main() {
  struct packet packets[CAP] = {};
  int begin = 0, queue_len = 0;

  while(1) {
    if (queue_len < CAP) {
      int end = (begin + queue_len)%CAP;
      if (receive_packet(&packets[end]) &&
          packets[end].port != 9)
        queue_len += 1;
    }
    if (0 < queue_len) {
      send_packet(&packets[begin]);
      begin = (begin + 1)%CAP;
    }
  }
  return 0;
}
\end{lstlisting}
  \caption{Original implementation of the Drop protocol. Contains an infinite
    loop and a latent ring implementation}
  \label{lst:orig}
\end{figure}

In order to absorb bursts, the code employs a ring buffer. The interface to the
network is: \code{bool receive_packet(struct packet* dst)}, which non-blockingly
reads an inbound packet and stores it at \code{dst}, returning success or
failure; \code{bool can_send_packet()}, which checks if a new packet can be
sent; and \code{void send_packet(struct packet* src)}, which sends the packet
stored at \code{src}. The application follows a common practice: it consists of a
main event loop, it has no other complex loops, and it uses a ring, a popular
data structure in network programming.

{\bf Step 1: Label infinite loops.} Our approach requires special handling of
infinite loops in the application code. A developer puts a label \code{LOOP}
into the condition in order to designate the potentially infinite loop. For a
typical NF there should be only a handful or them (1 in our example).

{\bf Step 2: Use library components. } A developer replaces every array with a
component from Vigor library that performs the same function. For our example,
the library provides a packet ring with a simple interface: \code{bool
  ring_empty(ring*)} checks whether there is a packet in the ring, \code{bool
  ring_full(ring*)} checks whether the ring has run out of space, \code{void
  ring_push_back(ring*, packet* src)} adds packet \code{src} to the ring, and
\code{void ring_pop_front(ring*, packet* dst)} removes a packet from the ring
and places it in \code{dst}. Fig.\ref{lst:vigor} shows the resulting NF code.
\code{VIGOR_CHECK} (line NN) adds additional property that we would like to
verify --- that we do not forward packets with port 9.

\begin{figure}[h!]
\begin{lstlisting}
int main(int argc, char** argv) {
  struct packet p;
  struct ring *r = ring_create(CAP);
  if (!r) return 1;
  while(LOOP(1))
  {
    loop_iteration_begin(&r);
    if (!ring_full(r))
      if (receive_packet(&p) && p.port != 9)
        ring_push_back(r, &p);
    if (!ring_empty(r) && can_send_packet()) {
      ring_pop_front(r, &p);
      VIGOR_CHECK(p.port != 9);
      send_packet(&p);
    }
    loop_iteration_end(&r);
  }
  return 0;
}
\end{lstlisting}
  \caption{Vigor implementation of Drop protocol. \code{LOOP} labels the
    infinite loop head and we explicitly use a ring from the Vigor library.
    \code{VIGOR_CHECK} at the line NN adds a semantic property to be verified.}
  \label{lst:vigor}
\end{figure}

@@

/*@
  fixpoint bool
  user_property(packet p) {
    switch(p) { case packet(port):
                  return port != 9; }
  }
  @*/

static bool
packet_constraints(struct packet* p) {
  return p->port != 9;
}
\end{lstlisting}
  \caption{Custom definitions capturing the desired property. They capture the
    essential property of every packet in the ring.}
  \label{lst:customizations}
\end{figure}

Currently, we have to duplicate the packet condition in two languages (formal
logic -lines NN, and C - NN) due to the limitations of our prototype.

{\bf Step 4: Run Vigor checker.} Vigor will find the appropriate symbolic model
and check that the application does not crash in any execution scenario. It will
then make sure that the chosen model is faithful to the formally proven
implementation of the component.

@@

\subsection{Under the hood}

@@

\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p)
/*@ requires ringp(r, ?user_property,
                   ?lst, ?cap) &*&
             lst != nil &*& packetp(p, _); @*/
/*@ ensures ringp(r, user_property,
                  tail(lst), cap) &*&
            packetp(p, head(lst)) &*&
            true == user_property(head(lst)); @*/
{
  //@ extract_first(r);
  struct packet* src_pkt = r->array + r->begin;
  p->port = src_pkt->port;
  r->len = r->len - 1;
  r->begin = r->begin + 1;
  if (r->cap <= r->begin) {
    r->begin = 0;
    //@ stitch_with_empty_overflow(r);
  } else {
    //@ stitch_with_empty(r);
  }
}
\end{lstlisting}
  \caption{A formal contract and verified implementation of the
    \code{ring_pop_front} function. The comments starting with \code{@} symbol
    are formal annotations that hold the function contract (\code{requires} -
    precondition, \code{ensures} - postcondition) and lemmas that help the proof
    checker to verify the state transformations in the code.}
  \label{lst:contract}
\end{figure}

@@

A model of \code{ring_pop_front} used for this example is shown on
Fig.~\ref{lst:symbolic-model}. This model is simple enough to enable
exhaustive SE, but preserves the vital propertiy --- \code{packet_constraints}
(Fig.\ref{lst:customizations}).

@@

Formal verification engine checks this sequence using the contracts like in the
Fig.~\ref{lst:contract}, and ensures the abscence of misuse e.g. call for
\code{ring_pop} on a possibly empty ring. It then checks that the model output
covers all the possible outputs specified in the contract for the tip function.
This happens in the last line on the listing. The assert contains the
path-constraint generated by the symbolic model. In this case the contract
(Fig.\ref{lst:contract}) guarantees that the popped packet satisfies
\code{user_property} that entails \code{arg2.port != 9}, hence the trace is
ligitimate. (see~\S\ref{sec:model-validation-criteria}).

>>>>

\lstset{language=C,
  escapechar=^,
  basicstyle=\footnotesize\ttfamily,
  keywordstyle=\color{myblue}\ttfamily,
  stringstyle=\color{mygreen}\ttfamily,
  commentstyle=\color{gray}\ttfamily,
  keepspaces=true,
  columns=flexible,
  tabsize=2,
  otherkeywords={bool},
  numbers=left,
  numbersep=1pt,
  numberstyle=\tiny\color{gray}
}

@@

This section motivates and explains at a high level what it's like to use Vigor
when developing NFs. For the sake of illustration, say we wish to develop a
trivial stateful NF that implements the discard protocol~\cite{rfc863}: an
infinite loop receives packets from one interface, discards the ones that target
port 9 (``discard port''), and forwards the others through another interface.
A possible implementation is on the Fig.~\ref{lst:orig}.

@@

int main() {
  struct packet packets[CAP] = {}; ^\label{line:pkt-def}^
  int begin = 0, queue_len = 0; ^\label{line:iter-def}^

  while(1) {^\label{line:loop-begin}^
    if (queue_len < CAP) {
      int end = (begin + queue_len)%CAP;
      if (receive_packet(&packets[end]) && ^\label{line:recv-pkt}^
          packets[end].port != 9)
        queue_len += 1;
    }
    if (0 < queue_len && can_send_packet()) { ^\label{line:can-send}^
      send_packet(&packets[begin]); ^\label{line:send-pkt}^
      begin = (begin + 1)%CAP;
    }
  }^\label{line:loop-end}^
  return 0;
}
\end{lstlisting}
  \caption{Original implementation of the Drop protocol. Contains an infinite
    loop (l.\ref{line:loop-begin}-\ref{line:loop-end}) and a latent ring
    implementation (declared on l.\ref{line:pkt-def},\ref{line:iter-def})}
  \label{lst:orig}
\end{figure}

In order to absorb bursts, the code employs a ring buffer. The application
interacts with the network using the following 3 functions:
\code{receive_packet}(l.\ref{line:recv-pkt}), which non-blockingly reads an
inbound packet and stores it at \code{packets[end]}, returning success or
failure; \code{can_send_packet}(l.\ref{line:can-send}), which checks if a new
packet can be sent; and \code{send_packet}(l.\ref{line:send-pkt}), which sends
the packet stored at \code{packets[begin]}. The application follows a common
practice: it consists of a main event loop, it has no other complex loops, and
it uses a ring, a popular data structure in network programming.

The ring keeps the state of our NF and makes many existing verification
approaches impractical.

Fig.~\ref{lst:vigor} shows the actual code that we are going to verify. The
first two steps show how we get there. \code{VIGOR_CHECK}
(l.~\ref{line:vigcheck}) adds additional property that we would like to verify
--- that we do not forward packets with port 9.

\begin{figure}[h!]
\begin{lstlisting}
int main(int argc, char** argv) {
  struct packet p;
  struct ring *r = ring_create(CAP);^\label{line:mk-ring}^
  if (!r) return 1;
  while(LOOP(1))^\label{line:loop-hd}^
  {
    loop_iteration_begin(&r);
    if (!ring_full(r))
      if (receive_packet(&p) && p.port != 9)
        ring_push_back(r, &p);
    if (!ring_empty(r) && can_send_packet()) {
      ring_pop_front(r, &p);
      VIGOR_CHECK(p.port != 9);^\label{line:vigcheck}^
      send_packet(&p);
    }
    loop_iteration_end(&r);
  }
  return 0;
}
\end{lstlisting}
  \caption{Vigor implementation of Drop protocol.
    \code{LOOP}(l.\ref{line:loop-hd}) labels the infinite loop head and we
    explicitly use a ring(l.\ref{line:mk-ring}) from the Vigor library. \code{VIGOR_CHECK} at the
    line \ref{line:vigcheck} adds a semantic property to be verified.}
  \label{lst:vigor}
\end{figure}

{\bf Step 1: Label infinite loops.} Our approach requires special handling of
infinite loops in the application code. A developer puts a label \code{LOOP}
(l.\label{line:loophd}) into the condition in order to designate the
potentially infinite loop. For a typical NF there should be only a handful or
them (1 in our example).

{\bf Step 2: Use library components. } A developer replaces every array with a
component from Vigor library that performs the same function. For our example,
the library provides a packet ring with a simple interface: \code{bool
  ring_empty(ring*)} checks whether there is a packet in the ring, \code{bool
  ring_full(ring*)} checks whether the ring has run out of space, \code{void
  ring_push_back(ring*, packet* src)} adds packet \code{src} to the ring, and
\code{void ring_pop_front(ring*, packet* dst)} removes a packet from the ring
and places it in \code{dst}.

@@

/*@^\label{line:vf-user-prop-from}^
  fixpoint bool
  user_property(packet p) {
    switch(p) { case packet(port):
                  return port != 9; }
  }
  @*/^\label{line:vf-user-prop-to}^

static bool^\label{line:c-user-prop-from}^
packet_constraints(struct packet* p) {
  return p->port != 9;
}^\label{line:c-user-prop-to}^
\end{lstlisting}
  \caption{Custom definitions capturing the desired property. They capture the
    essence of every packet in the ring. Currently, we have to duplicate the
    packet condition in two languages (formal logic --
    ll\ref{line:vf-user-prop-from}-\ref{line:vf-user-prop-to}, and C --
    ll\ref{line:c-user-prop-from}-\ref{line:c-user-prop-to}) due to the
    limitations of our prototype.}
  \label{lst:customizations}
\end{figure}

{\bf Step 4: Run Vigor checker.} Vigor will find the appropriate symbolic model
and check that the application does not crash in any execution scenario. It will
then make sure that the chosen model is faithful to the formally proven
implementation of the component. \S\ref{sec:under-the-hood} provides more
details on how it is done.

@@

\subsection{Under the hood}
\label{sec:under-the-hood}

@@

\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p)
/*@ requires ringp(r, ?user_property,^\label{line:precondition-from}^
                   ?lst, ?cap) &*&
             lst != nil &*& packetp(p, _); @*/^\label{line:precondition-to}^
/*@ ensures ringp(r, user_property,^\label{line:postcondition-from}^
                  tail(lst), cap) &*&
            packetp(p, head(lst)) &*&
            true == user_property(head(lst)); @*/^\label{line:postcondition-to}^
{
  //@ extract_first(r); ^\label{line:lemma1}^
  struct packet* src_pkt = r->array + r->begin;
  p->port = src_pkt->port;
  r->len = r->len - 1;
  r->begin = r->begin + 1;
  if (r->cap <= r->begin) {
    r->begin = 0;
    //@ stitch_with_empty_overflow(r);^\label{line:lemma2}^
  } else {
    //@ stitch_with_empty(r);^\label{line:lemma3}^
  }
}
\end{lstlisting}
  \caption{A formal contract and verified implementation of the
    \code{ring_pop_front} function. The comments starting with \code{@} symbol
    are formal annotations that hold the function contract
    (\code{requires}(ll.\ref{line:precondition-from}-\ref{line:precondition-to})
    - precondition,
    \code{ensures}(ll.\ref{line:postcondition-from}-\ref{line:postcondition-to})
    - postcondition) and
    lemmas(ll.\ref{line:lemma1},\ref{line:lemma2},\ref{line:lemma3}) that help
    the proof checker to verify the state transformations in the code.}
  \label{lst:contract}
\end{figure}

@@

A model of \code{ring_pop_front} used for this example is shown on
Fig.~\ref{lst:symbolic-model}. This model is simple enough to enable
exhaustive SE, but preserves the vital property --- \code{packet_constraints}
(Fig.\ref{lst:customizations}).

@@

Formal verification engine checks this sequence using the contracts like in the
Fig.~\ref{lst:contract}, and ensures the abscence of misuse e.g. call for
\code{ring_pop} on a possibly empty ring. It then checks that the model output
covers all the possible outputs specified in the contract for the tip function.
This happens in the last line on the listing. The assert contains the
path-constraint generated by the symbolic model. In this case the contract
(Fig.\ref{lst:contract}) guarantees that the popped packet satisfies
\code{user_property} that entails \code{arg2.port != 9}, hence the trace is
legitimate. (see~\S\ref{sec:model-validation-criteria}).

\subsection{Misfit symptoms}
